For submission to the Journal of Integrated Pest Management

### 2021-09-21 ofv_fife_2021.Rmd
updated section title, it doesn't reflect a description of the mite, just describes relationships
Updated to reflect editorial comments

---
title: "For submission to the Journal of Integrated Pest Management"
subtitle: "Category: 'Issues'"
output: 
  word_document:
    reference_docx: ../r-word-templates/word-template-jis.docx
# - html_document
bibliography: 'bib/thesis.bib'
citation_package: natbib
biblio-style: apalike
csl: 'entsoc.csl'
#always_allow_html: true
---
```{r setup, include=FALSE}
#setup options
knitr::opts_chunk$set(fig.width=12, fig.height=8, echo=FALSE)

@@

#### Mite Description

Mite taxonomy is complicated by cryptic species complexes which occur in many plant-feeding groups of the Acari [@Arthur2011; @Skoracka2010; @Skoracka2013; @Umina1999], including tenuipalpid mites from the genus *Brevipalpus* [@Navia2013]. The commonly used phase-contrast microscopy is insufficient to detect some diagnostic characters for separation of cryptic species, instead best practices recommend the combination of Differential Interference Contrast (DIC) Microscopy and Scanning Electron Microscopy along with molecular methods to separate cryptic species [@Beard2015]. The flat mites collected were initially suspected to belong to *B. californicus* after inspection with phase contrast microscopy. Subsequent observation via DIC microscopy at FDACS agreed with this tentative identification. Unfortunately, the *B. californicus* s.l. species group, *sensu* @Baker1987 is suspected to contain cryptic species [@Childers2011; @Rodrigues2013]. New mite samples were collected from symptomatic liriopogons and *A. elatior* in Leon County and sent to USDA-ARS's Electron and Confocal Microscopy Unit for analysis. Three mite species were recovered and examined under cryo-scanning electron microscopy (Cryo-SEM): *B. californicus* s.l. (`r citef("fig_3")`), *B. obovatus* Donnadieu and *B. confusus* Baker. The recent report of OFV in the US is thought to be @Ko1985 which describes nuclear inclusions caused by an undescribed bacilliform rhabdovirus in *Brassia* orchids. The significance of this report is their description of the spoke-wheel configurations of the viral particles [@Ko1985], a sign typically associated with OFV infection [@Chang1976]. Unfortunately, this article made no mention of mites or further investigations of the virus. The first report of OFV in the continental US was @Bratsch2015, who confirmed the presence of OFV in *Phalaenopsis* hybrids using Transmission Electron Microscopy of ultrathin sections of plant tissue as well as molecular sequence analysis. They also discuss the association of OFV with *Brevipalpus* mites, but the authors did not make a conclusive species identification beyond suggesting that the mite vector belonged to the *B. californicus* group, referring to @Kondo2003's publication [@Bratsch2015]. Later reports of OFV described OFV infecting a previously undescribed Nolinoidaea hosts in Australia [@Dietzgen2018b; @Mei2016], including *Liriope spicata* (Thunb.) Lour, a different species of liriopogon than those identified from the Florida sites. We are not aware of any reports of OFV infecting liriopogons, *A. elatior* nor other Nolinoidaea in the US. Although @Peng2013 had mentioned an association between *B. californicus* and *A. elatior*, they never reported symptoms of OFV-Orc in this plant. We believe that our findings indicate the first report of OFV-Orc infecting ornamental Nolinoidaea in Florida, and possibly the US. This publication also marks the first reports of *A. elatior* and *Ophiopogon* spp. as natural hosts of OFV-Orc. There are two orchid strains of OFV (OFV-Orc1 and OFV-Orc2), and two citrus strains (OFV-Cit1 and OFV-Cit2) [@Roy2020; @Beltran-Beltran2020]. The OFV strains detected in Florida are identical in genome sequence to the orchid strains of OFV infecting citrus in Hawaii, Mexico, Colombia, and South Africa [@Roy2020; @Beltran-Beltran2020]. Both OFV-Orc1 and OFV-Orc2 infect citrus [@Roy2020], but none of the citrus  strains have been reported from any orchid species. The *Brevipalpus* mites collected from liriopogons and *A. elatior* in Leon County were abundant on OFV-infected plants very near to citrus trees, some plants even surrounding the trunk. *B. californicus* s. l. has been reported as a pest of citrus [@Childers2003] and are often collected from citrus fruits [@Vacante2016; @Vacante2010; @Baker1987; @Baker1949]. The proximity of these mite vectors to citrus raises the question: why these trees are not currently infected with OFV-Orc? It is important to note the uncertainty surrounding the vector for OFV-Orc. There are three mite species which have been recovered from OFV-Orc infected plants: *B. obovatus*, and *B. confusus* and *B. californicus* s.l., but only *B. californicus* has been described as a vector of OFV. Even so, the *B. californicus* which we find on liriopogons and *A. elatior* may not be the same cryptic species as those found on citrus. Transmission of OFV from populations of *B. californicus* liriopogon/*A. elatior* to citrus may be limited by host preferences, vectorial capacity, viral propagation/circulation in the vector, viral acquisition times, or feeding times required for transmission to citrus. Even so, these types of questions require future study to determine the potential of nolinoidaea to citrus transmission. Best practices for integrated pest management have not been created for controlling *Brevipalpus* mites on these ornamentals, but methods designed to control *Brevipalpus* in other systems may be applicable. The most common method used to control *Bervipalpus* are synthetic acaricides [@Andrade2010; @Andrade2019]. Unfortunately, some acaricides and their residues can harm beneficial predatory mites as well [@Fernandez2017], even at low doses [@Havasi2021], and mixing different chemistries can be detrimental for mite control [@Vechia2018]. In addition, pesticide resistance has been reported in various *Brevipalpus* populations [@Rocha2021; @Alves2000; @Campos2002; @Omoto2000], due to exposure to pesticides used to control other arthropod pests [@Vechia2021]. In addition, predatory mites [@Argolo2020; @Chen2006], entomopathogenic fungi [@RossiZalaf2008; @Magalhaes2005; @Pena2015; @Revynthi2019] have shown promise for controlling other *Brevipalpus* mites. Moreover, it is often possible to integrate different control techniques for improved management, such as combining predatory mites with compatible acaricides and entomopathogenic fungi [@Reddy2001; @Midthassel2016; @Andrade2019]. In conclusion, detecting OFV in Florida represents a concern for horticulturists who grow orchids, *Liriope*, *Ophiopogon*, or other susceptible Asparagaceae species which are commonly used in landscaping. Florida is also home to a plethora of native and naturalized orchid species, many of which are threatened, including cultivated *Vanilla* in southern Florida [@Chambers2019] and the famous Ghost Orchid, [*Dendrophylax lindenii* (Lindl.) Benth. ex Rolfe]. Citrus leprosis was present in Florida during the 1860’s and almost eradicated by the mid-1960s [@Childers2003; @Knorr1968a; @Knorr1968b]. An examination of herbarium specimens of Florida citrus found that this historical virus, Citrus leprosis dichorhavirus-N0, is distantly related to the modern isolates of OFV [@Hartung2015; @Kitajima2011a; @Roy2020]. The recent detection of OFV-Orc1 in South Africa [@Cook2019] in *C. sinensis* (Navel and Valencia orange) and OFV-Orc2 in Hawaii [@Velarde2021] in *C. reticulata* (mandarin) and *C. jambhiri* (rough lemon) associated with leprosis-like symptoms highlights the potential threat of different isolates of OFV on citrus, which will be a definite concern to the US multi-billion-dollar citrus industry already impacted by the Huanglongbing disease. *B. californicus*, *B. yothersi*, and *B. obovatus* are all present in Florida [@Childers2003; @Akyazi2017], and are difficult to identify by non-experts, or without advanced methodologies. DNA barcoding [@Armstrong2005] or a similarly simple and accurate method for identification of these mite complexes is vital to identify mite populations which need to be monitored or controlled. By doing so, we can determine the risk OFV-Orc represents for the native plants, agriculture and the ornamental/landscaping industries of Florida and the surrounding regions.

>>>>

---
title: "For submission to the Journal of Integrated Pest Management"
subtitle: "Category: 'Issues'"
output: 
  word_document:
    reference_docx: word-template-jis.docx
# - html_document
bibliography: 'bib/thesis.bib'
citation_package: natbib
biblio-style: apalike
csl: 'entsoc.csl'
#always_allow_html: true
---
```{r setup, include=FALSE}
#setup options
knitr::opts_chunk$set(fig.width=12, fig.height=8, echo=FALSE)

@@

#### A comment on the the status of *Brevipalpus* in Florida

Mite taxonomy is complicated by cryptic species complexes which occur in many plant-feeding groups of the Acari [@Arthur2011; @Skoracka2010; @Skoracka2013; @Umina1999], including tenuipalpid mites from the genus *Brevipalpus* [@Navia2013]. The commonly used phase-contrast microscopy is insufficient to detect some diagnostic characters for separation of cryptic species, instead best practices recommend the combination of Differential Interference Contrast (DIC) Microscopy and Scanning Electron Microscopy along with molecular methods to separate cryptic species [@Beard2015]. The flat mites collected were initially suspected to belong to *B. californicus* after inspection with phase contrast microscopy. Subsequent observation via DIC microscopy at FDACS agreed with this tentative identification. Unfortunately, the *B. californicus* s.l. species group, *sensu* @Baker1987 is suspected to contain cryptic species [@Childers2011; @Rodrigues2013]. New mite samples were collected from symptomatic liriopogons and *A. elatior* in Leon County and sent to USDA-ARS's Electron and Confocal Microscopy Unit for analysis. Three mite species were recovered and examined under cryo-scanning electron microscopy (Cryo-SEM): *B. californicus* s.l. (`r citef("fig_3")`), *B. obovatus* Donnadieu and *B. confusus* Baker. The recent report of OFV in the US is thought to be @Ko1985 which describes nuclear inclusions caused by an undescribed bacilliform rhabdovirus in *Brassia* orchids. The significance of this report is their description of the spoke-wheel configurations of the viral particles [@Ko1985], a sign typically associated with OFV infection [@Chang1976]. Unfortunately, this article made no mention of mites or further investigations of the virus. The first report of OFV in the continental US was @Bratsch2015, who confirmed the presence of OFV in *Phalaenopsis* hybrids using Transmission Electron Microscopy of ultrathin sections of plant tissue as well as molecular sequence analysis. They also discuss the association of OFV with *Brevipalpus* mites, but the authors did not make a conclusive species identification beyond suggesting that the mite vector belonged to the *B. californicus* group, referring to @Kondo2003's publication [@Bratsch2015]. Later reports of OFV described OFV infecting a previously undescribed Nolinoidaea hosts in Australia [@Dietzgen2018b; @Mei2016], including *Liriope spicata* (Thunb.) Lour, a different species of liriopogon than those identified from the Florida sites. We are not aware of any reports of OFV infecting liriopogons, *A. elatior* nor other Nolinoidaea in the US. Although @Peng2013 had mentioned an association between *B. californicus* and *A. elatior*, they never reported symptoms of OFV-Orc in this plant. We believe that our findings indicate the first report of OFV-Orc infecting ornamental Nolinoidaea in Florida, and possibly the US. This publication also marks the first reports of *A. elatior* and *Ophiopogon* spp. as natural hosts of OFV-Orc. There are two orchid strains of OFV (OFV-Orc1 and OFV-Orc2), and two citrus strains (OFV-Cit1 and OFV-Cit2) [@Roy2020; @Beltran-Beltran2020]. The OFV strains detected in Florida are identical in genome sequence to the orchid strains of OFV infecting citrus in Hawaii, Mexico, Colombia, and South Africa [@Roy2020; @Beltran-Beltran2020]. Both OFV-Orc1 and OFV-Orc2 infect citrus [@Roy2020], but none of the citrus  strains have been reported from any orchid species. It is important to note the uncertainty surrounding the vector for OFV-Orc. There are three mite species which have been recovered from OFV-Orc infected plants: *B. obovatus*, and *B. confusus* and *B. californicus* s.l., but only *B. californicus* has been described as a vector of OFV. Even so, these types of questions require future study to determine the potential of nolinoidaea to citrus transmission. Best practices for integrated pest management have not been created for controlling *Brevipalpus* mites on these ornamentals, but methods designed to control *Brevipalpus* in other systems may be applicable. The most common method used to control *Bervipalpus* are synthetic acaricides [@Andrade2010; @Andrade2019]. Unfortunately, some acaricides and their residues can harm beneficial predatory mites as well [@Fernandez2017], even at low doses [@Havasi2021], and mixing different chemistries can be detrimental for mite control [@Vechia2018]. In addition, pesticide resistance has been reported in various *Brevipalpus* populations [@Rocha2021; @Alves2000; @Campos2002; @Omoto2000], due to exposure to pesticides used to control other arthropod pests [@Vechia2021]. In addition, predatory mites [@Argolo2020; @Chen2006], entomopathogenic fungi [@RossiZalaf2008; @Magalhaes2005; @Pena2015; @Revynthi2019] have shown promise for controlling other *Brevipalpus* mites. Moreover, it is often possible to integrate different control techniques for improved management, such as combining predatory mites with compatible acaricides and entomopathogenic fungi [@Reddy2001; @Midthassel2016; @Andrade2019]. In conclusion, detecting OFV in Florida represents a concern for horticulturists who grow orchids, *Liriope*, *Ophiopogon*, or other susceptible Asparagaceae species which are commonly used in landscaping. Florida is also home to a plethora of native and naturalized orchid species, many of which are threatened, including cultivated *Vanilla* in southern Florida [@Chambers2019] and the famous Ghost Orchid, [*Dendrophylax lindenii* (Lindl.) Benth. ex Rolfe]. Citrus leprosis was present in Florida during the 1860’s and almost eradicated by the mid-1960s [@Childers2003; @Knorr1968a; @Knorr1968b]. An examination of herbarium specimens of Florida citrus found that this historical virus, Citrus leprosis dichorhavirus-N0, is distantly related to the modern isolates of OFV [@Hartung2015; @Kitajima2011a; @Roy2020]. The recent detection of OFV-Orc1 in South Africa [@Cook2019] in *C. sinensis* (Navel and Valencia orange) and OFV-Orc2 in Hawaii [@Velarde2021] in *C. reticulata* (mandarin) and *C. jambhiri* (rough lemon) associated with leprosis-like symptoms highlights the potential threat of different isolates of OFV on citrus. *B. californicus*, *B. yothersi*, and *B. obovatus* are all present in Florida [@Childers2003; @Akyazi2017], and are difficult to identify by non-experts, or without advanced methodologies. DNA barcoding [@Armstrong2005] or a similarly simple and accurate method for identification of these mite complexes is vital to identify mite populations which need to be monitored or controlled. By doing so, we can determine the risk OFV-Orc represents for the native plants, agriculture and the ornamental/landscaping industries of Florida and the surrounding regions.
